---
title: "Getting started with GenSEM"
author: "Carlos Aloisio (caloisio@gmail.com)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with GenSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

GenSEM fits multilevel multiprocess hazard systems: several event processes
observed on the same people, whose unobserved person-level determinants are
allowed to be correlated. The recursive class handled here pairs a discrete
selection equation (who gets the exposure) with a duration outcome (how long
until the event), estimated jointly by Gauss-Hermite quadrature so that
selection on unobservables does not bias the exposure effect.

This vignette simulates a small system, fits it, and reads the output. It
runs in a few seconds and uses no external data.

## A simulated selection-plus-hazard system

We simulate `n` clusters (say, mothers). A latent trait `u` raises the death
hazard; a correlated trait `v` raises the chance of treatment. Ignoring that
correlation is exactly what biases naive hazard regressions.

```{r}
library(GenSEM)
set.seed(42)
n <- 120
u <- rnorm(n, 0, 0.6)                      # hazard-side frailty
v <- 0.5 * u + rnorm(n, 0, 0.5)            # correlated selection trait
sel <- data.frame(id = 1:n, z = rnorm(n))  # z: instrument (selection only)
sel$treat <- as.integer(0.8 * sel$z - 0.3 + v > 0)
haz <- data.frame(id = 1:n, treat = sel$treat, dur = rexp(n, 1) + 0.1)
haz$death <- rbinom(n, 1, plogis(-0.8 - 0.6 * haz$treat + u))
```

## Joint fit

The selection equation gets the instrument `z`; the hazard equation gets the
endogenous `treat`. `exposure` names the time-at-risk column.

```{r}
fit <- fit_recursive(treat ~ z, death ~ treat,
                     data_sel = sel, data_haz = haz,
                     id = "id", exposure = "dur", Q = 9)
summary(fit)
```

```{r}
confint(fit, "haz:treat")
```

## Reading the output

The `haz:treat` row is the treatment effect on the log-hazard net of
selection on unobservables: the joint model uses the estimated covariance
between the two random effects (the `Sigma` block) to absorb the part of the
treatment-outcome association that is due to shared latent traits. A naive
one-equation Poisson fit of the same data attributes that shared part to the
treatment itself. The `sel:` rows are the probit selection equation, where
the instrument `z` should be strong; and `Sigma` reports the estimated
variances of `u` and `v` and their covariance, whose sign tells you the
direction of the selection.

## Scope note: sampling weights

GenSEM has **no support for sampling weights** (nor PSU/strata): standard
errors are model-based and estimation is unweighted maximum likelihood. For
complex-survey data, handle the design through covariates plus sensitivity
analyses, or use a design-based tool. Weighted pseudo-likelihood is on the
roadmap but is not available in this version.
